---
title: "Create life history object"
author: "Jason Doll"
date: "`r Sys.Date()`"
format: 
  html: 
    code-fold: false
    toc: true
    toc-depth: 2
    reference-location: margin
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
    echo: true
    fig-width: 7.5
    fig-height: 4.5
    fig-align: center
---

```{r}
#| label: setup
#| results: hide
#| message: false
library(rFAMS)
library(FSA)     ## for growth model 
library(FSAdata) ## for data used in example
```

The objective of this article is to demonstrate how to build a life history object that is used in all of rFAMS core functions (yield-per-recruit modeling [`yprBH_minLL_fixed()`, `yprBH_minLL_var()`, `yprBH_SlotLL()`] and dynamic pool modeling [`dpmBH()`].

# Build a life history parameter object from user specified values

Life history parameters required for yield-per-recruit and dynamic pool modeling include the initial number of recruits, maximum age, von Bertalanffy growth model parameters ($L_\infty$, $K$, and $t_0$) and parameters from the log10-transformed weight length model (alpha and beta). These must be stored in a list or vector, by default `makeLH()` returns values in list. Users have the option of building a list manually but the `makeLH()` function contains a series of checks that can identify potential errors. 

If a list is returned then values will be displayed with the number of decimals provided by the user. If a vector is returned then the number of decimals displayed will be the same for each value and will match the value supplied by the user with the most decimals. Thus, a list is preferred as it will be easier to match what was given to what was expected to be given.

The first example explicitly specifies each argument. The two created objects below (LHparamsList and LHparamsVector) can then be used in any of the core rFAMS functions.

```{r}
#| label: Specify arguments in life history object
#Return parameters as a list
LHparamsList <- makeLH(N0=100,tmax=15,Linf=500,K=0.3,t0=-0.5,LWalpha=-5.613,LWbeta=3.1)
#View life history object
LHparamsList

#Return parameters as a vector
LHparamsVector <- makeLH(N0=100,tmax=15,Linf=500,K=0.3,t0=-0.5,LWalpha=-5.613,LWbeta=3.1,
                         restype="vector")
#View life history object
LHparamsVector
```

# Build a life history parameter object from model output

The `makeLH()` function also allows model objects as input for von Bertalanffy growth and length-weight models. See the [FSA package](https://fishr-core-team.github.io/FSA/index.html) for more information about fitting von Bertalanffy growth models and length-weight models.

```{r}
#| label: Generate model objects for life history
## Generate LVB results using the SptoVA1 data from the FSA package
data(SpotVA1,package="FSA")
SpotVA1 <- SpotVA1 |>
  dplyr::mutate(tl=tl*25.4)
vb1 <- FSA::makeGrowthFun(type="von Bertalanffy")
fit1 <- nls(tl~vb1(age,Linf,K,t0),data=SpotVA1,
            start=FSA::findGrowthStarts(tl~age,data=SpotVA1))

## Generate length-weight regression results using the BluegillLM data from the FSAdata package
data(BluegillLM,package="FSAdata")
BluegillLM <- BluegillLM |>
  dplyr::mutate(logW=log10(wght),
                logL=log10(tl))
fit2 <- lm(logW~logL,data=BluegillLM)
```

The model object created above are then passed through the `makeLH()` function in the `Linf` and `LWalpha` arguments.

```{r}
#| label: Use model objects in life history object
LHparams <- makeLH(N0=100,tmax=15,Linf=fit1,LWalpha=fit2)
LHparams
```

