---
title: "Weight-Length"
date: "`r Sys.Date()`"
format: 
  html: 
    code-fold: false
    toc: true
    toc-depth: 2
    reference-location: margin
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
    echo: true
    fig-width: 4
    fig-height: 4
    fig-align: center
---

```{r}
#| echo: false
#| eval: false
# Use this to keep up with development version ... eventually delete
remotes::install_github('fishR-Core-Team/rFAMS',ref="dev")
```

::: {.callout-warning}
This is a work-in-progress.
:::

# Introduction
The primary response or dependent variable calculated within the yield-per-recruit (YPR) and dynamic pool (DPM) models implemented in `rFAMS` is yield. *Yield* is the amount of fish harvested from a fishery over a specified period of time and is usually, as is done in `rFAMS`, expressed in units of weight or biomass (e.g., g, kg, tons). As such, the YPR and DPM models need a mechanism to model weight of individual or "average" fish. This is accomplished through the *weight-length relationship*. The purpose of this article is to briefly describe the typical characteristics of weight-length relationships for fish, briefly show how to model a weight-length relationship in `R`, and show how to extract the necessary weight-length relationship parameters for use in the YPR and DPM models in `rFAMS`.

The following packages are used in this article.

```{r}
#| results: hide
#| message: false
library(rFAMS)
library(ggplot2)
library(ggtext)   # for element_markdown()
```

# Weight-Length Relationship
The required data for examining the weight-length relationship for a sample of fish is measurements of the weight and length of individual fish at the time of capture. **In `rFAMS`, weight is assumed to be recorded in g and length in mm.** Other data may be recorded as well (e.g., year of capture, area of capture, sex), but these other variables are not be considered in this article.

The relationship between fish weight and length tends to have two important characteristics, as illustrated for several disparate species in @fig-WL_4species. First, the relationship is not linear. The degree of non-linearity may be impacted by the range of lengths observed, with narrower ranges appearing more linear (e.g., compare Walleye to Bluegill in @fig-WL_4species). This non-linearity appears as length is inherently a one-dimensional meausure whereas weight is more akin to the three-dimensional measure of volume. Second, the variability in weight at a given length increases as the length of the fish increases (i.e., the vertical scatter of the points increases from left-to-right in @fig-WL_4species). Some of this difference is due to scale (i.e., less room for variability at smaller values of length), but some is also due to the natural increase in variability among fish over time (e.g., due to differences in foot intake and energy expenditures).

```{r}
#| label: prepdata_4species
#| echo: false
#| results: hide
data(BlackDrum2001,package="FSAdata")
BlackDrum2001 <- BlackDrum2001 |>
  dplyr::select(spname,weight,tl) |>
  dplyr::rename(species=spname,w=weight) |>
  dplyr::mutate(w=w*454)
data(SculpinALTER,package="FSAdata")
SculpinALTER <- SculpinALTER |>
  dplyr::mutate(species="Slimy Sculpin") |>
  dplyr::select(species,w,tl)
data(WalleyeErie2,package="FSAdata")
WalleyeErie2 <- WalleyeErie2 |>
  dplyr::filter(loc==3,year==2010) |>
  dplyr::mutate(species="Walleye") |>
  dplyr::select(species,w,tl)
data(InchLake2,package="FSAdata")
BluegillIL <- InchLake2 |> 
  dplyr::filter(species=="Bluegill") |>
  dplyr::select(species,weight,length) |>
  dplyr::rename(w=weight,tl=length) |>
  dplyr::mutate(tl=tl*25.4)

dat <- dplyr::bind_rows(BlackDrum2001,SculpinALTER,WalleyeErie2,BluegillIL) |>
  dplyr::mutate(species=factor(species,
                               levels=c("Slimy Sculpin","Bluegill",
                                        "Walleye","Black Drum")),
                logw=log10(w),logl=log10(tl))
dat <- dat[complete.cases(dat),]
```

```{r}
#| label: fig-WL_4species
#| echo: false
#| fig-cap: Weight-length relationship for four disparate species.
#| fig-width: 7
#| fig-height: 7
ggplot(data=dat,mapping=aes(x=tl,y=w)) +
  geom_point(alpha=0.5) +
  scale_y_continuous(name="Weight (g)",limits=c(0,NA)) +
  scale_x_continuous(name="Total Length (mm)") +
  theme_bw() +
  facet_wrap(vars(species),scales="free")
```

These characteristics of weight-length data suggest that a two-parameter power function (to capture the curvature) with a multiplicative error term (to capture the increase in variance) should be used to model the weight-length relationship. Specifically, the model typically used with fish is

$$ W_i=aL_i^b\epsilon_i $$  {#eq-WLraw}

where $W_i$ and $L_i$ are the observed weight and length of the $i$th individual, $a$ and $b$ are constants to be estimated (i.e., model parameters), and $\epsilon_i$ is the multiplicative error term for the $i$th fish. The weight-length model in @eq-WLraw can be transformed to a linear model by taking the logarithms of both sides and simplifying; i.e.,

$$ log_{10}\large(W_i\large)=log_{10}(a)+blog_{10}\large(L_i\large)+log_{10}\large(\epsilon_i\large) $$  {#eq-WLtrans}

Any base of logarithms will work, but **`rFAMS` assumes that common logs ($log_{10}$) are used**. 

With this transformation, @eq-WLtrans is a linear function with $y=log_{10}\large(W_i\large)$, $x=log_{10}\large(L_i\large)$, slope=$b$, and y-intercept=$log_{10}(a)$. In addition, the individual errors (i.e., $log_{10}\large(\epsilon_i\large)$) in @eq-WLtrans are additive rather than multiplicative, such that the variability in $log_{10}\large(W_i\large)$ at a given $log_{10}\large(L_i\large)$ is constant for all $log_{10}\large(W_i\large)$. Thus, the log-log transformation for weight and length data tends to produce a relationship that is linear with a constant variability, as illustrated in @fig-logWL_4species for the same four disparate species.

```{r}
#| label: fig-logWL_4species
#| echo: false
#| fig-cap: Log10-transformed weight-length relationship for four disparate species.
#| fig-width: 7
#| fig-height: 7
ggplot(data=dat,mapping=aes(x=logl,y=logw)) +
  geom_point(alpha=0.5) +
  scale_y_continuous(name="log~10~ Weight (g)") +
  scale_x_continuous(name="log~10~ Total Length (mm)") +
  theme_bw() +
  theme(axis.title.x=ggtext::element_markdown(),
        axis.title.y=ggtext::element_markdown()) +
  facet_wrap(vars(species),scales="free")
```

Transformed weight-length data tend to be very strongly correlated, with $r^2$ values often in excess of 0.95. Individual fish that are clearly "off" the linear relationship evident for the rest of the sample often have measurements that are in error. For example, the Slimy Sculpin in @fig-logWL_4species at a log length of ~1.6 and log weight of ~0 very likely has a length measurement that is too low or a weight that is too high. Another common problem with smaller individuals of some species is that weight may have been taken with an instrument that lacked suitable precision for small individuals. This will appear in transformed weight-length plots as higher variability in log weight at lower log length values. This is apparent for Bluegill in @fig-logWL_4species. Individuals that are in error should either be corrected or removed from further analysis, and individuals for which the weight was measured imprecisely may also be removed.

The exponent parameter ($b$) of the weight-length relationship averages to approximately 3 across many, many species of fish (@fig-LWb). For most fish, $b$ is between 2.5 and 3.5 and values of $b$ less than 2.0 or greater than 4.0 should be reconsidered as suspect. It will be shown below how `rFAMS` uses $b$, but take note here that `rFAMS` will issue a warning if you attempt to use values of $b$ outside the range of 2 to 4.

```{r}
#| label: fig-LWb
#| echo: false
#| warning: false
#| message: false
#| fig-cap: Histogram of $b$ values from weight-length relationships for all relationships in the `rfishbase` database.
#| fig-height: 3.5
#| cache: true
# !!! I set cache: true as this takes many seconds to access rfishbase
# !!! It requires having rfishbase package installed
tmp <- rfishbase::length_weight(server="fishbase")
ggplot(data=tmp,mapping=aes(x=b)) +
  geom_histogram(color="black",fill="gray50") +
  scale_x_continuous(limits=c(1.75,4.1)) +
  scale_y_continuous(name="Frequency",expand=expansion(mult=c(0,0.05))) +
  theme_bw()
```


# Modeling Weight-Length in R
The advantage of log-log transforming weight-length data is that typical simple linear regression tools can be used to model the relationship. In this section, we show how to use `R` to model this linear relationship. This description assumes that you are familiar with the basic principles of simple linear regression.

The weight and length of Walleye from Lake Erie stored in `WalleyeErie2` in the `FSAdata` package will be used as an example. These data are obtained and the first few rows examined below. Note that weights and lengths are in `w` and `tl`, respectively.^[More information about these data can be found in [the documentation](https://fishr-core-team.github.io/FSAdata/reference/WalleyeErie2.html) for `WalleyeErie2`.]

```{r}
data(WalleyeErie2,package="FSAdata")
head(WalleyeErie2)
```

For our purposes, fish from one location (i.e., 3) and year (i.e,. 2010) will be isolated using `filter()` from `dplyr`.^[These are the "Walleye" data in fig-WL_4species.] These data are stored in the new data.frame `waeredux`.

```{r}
waeredux <- WalleyeErie2 |>
  dplyr::filter(loc==3,year==2010)
```

Transformed versions of the weight and length variables are added to the data.frame with `mutate()` from `dplyr`. Note that `log10()` computes common (i.e., base 10) logarithms.^[`log()` is used for natural (i.e., base $e$) logarithms.]

```{r}
waeredux <- waeredux |>
  dplyr::mutate(logw=log10(w),
                logtl=log10(tl))
head(waeredux)
```

A simple linear regression model is fit with `lm()` using a formula of the form `Y~X` as the first argument and the corresponding data.frame in `data=`. The result should be saved to an object for further analysis.

```{r}
waelm <- lm(logw~logtl,data=waeredux)
```

The estimates of $log_{10}(a)$ and $b$ are extracted from the saved `lm()` object with `coef()`. The estimate of $log_{10}(a)$ is under `(Intercept)` and the estimates of $b$ is under the name of the "X" variable (here `logtl`). Thus, the estimate of $log_{10}(a)$ is `r formatC(coef(waelm)[[1]],format="f",digits=6)` and the estimate of $b$ is `r formatC(coef(waelm)[[2]],format="f",digits=6)`.

```{r}
coef(waelm)
```

Simple linear regression analyses may require further work than what is shown here -- such as assumption checking, assessing adequacy of fit, testing significance. More detailed explanations of simple linear regression should be consulted for these other aspects. However, it is good practice to examine the fitted model relative to the observed data (@fig-fitPlot) to identify any gross problems with the data or model fitting. Below demonstrates one way to do this with `ggplot2`.^[[Here](https://fishr-core-team.github.io/fishR/blog/posts/2023-3-31_Milleretal2022_Fig23/#recreating-figure-3) is an example for modifications to this simple plot.]

```{r}
#| label: fig-fitPlot
#| fig-cap: Log10-transformed weight-length relationship for Walleye captured at location 3 in Lake Erie in 2010. The best-fit linear model is shown in blue.
#| message: false
ggplot(data=waeredux,mapping=aes(x=logtl,y=logw)) +
  geom_point(alpha=0.5) +
  geom_smooth(method="lm") +
  scale_y_continuous(name="log~10~ Weight (g)") +
  scale_x_continuous(name="log~10~ Total Length (mm)") +
  theme_bw() +
  theme(axis.title.x=ggtext::element_markdown(),
        axis.title.y=ggtext::element_markdown())
```

# Extracting Parameters for Use in `rFAMS`
Functions to perform the YPR and DPM modeling in `rFAMS` all take a list or vector that contains seven required life history parameters in the `lhparms=` argument^[See [here](https://fishr-core-team.github.io/rFAMS/reference/yprBH_MinLL.html) and [here](https://fishr-core-team.github.io/rFAMS/reference/dpmBH_MinLL.html), for example]. Two of those required life history parameters are $log_{10}(a)$ called `LWalpha` in `rFAMS` and $b$ called `LWbeta` in `rFAMS`. 

`makeLH()` is a convenience function in `rFAMS` that takes user-provided values for the seven life history parameters, performs adequacy checks on each,^[For example, is it numeric or is it >0 (if appropriate).] and then puts the values into a properly formatted list (preferably) or vector.^[See [`makeLH()` documentation](https://fishr-core-team.github.io/rFAMS/reference/makeLH.html) for more details.] In its simplest usage, `makeLH()` has seven required arguments, one for each of the required life history parameters. Two of these arguments are `LWalpha=` and `LWbeta=` for the two weight-length relationship parameters estimated in the previous section.

```{r}
LH <- makeLH(N0=100,tmax=15,Linf=500,K=0.3,t0=-0.5,
             LWalpha=-5.877308,LWbeta=3.341721)
LH
```

A less prone-to-error method for entering the weight-length relationship parameters is to give the object saved from `lm()` in the previous section to `LWalpha=` and not provide a value to `LWbeta=`. `makeLH()` will extract the parameter estimates from the `lm()` object to put in the life history parameter list.

```{r}
LH <- makeLH(N0=100,tmax=15,Linf=500,K=0.3,t0=-0.5,
             LWalpha=waelm)
LH
```

