---
title: "Estimate yield based on multiple minimum length limits"
author: "Jason Doll"
date: "`r Sys.Date()`"
format: 
  html: 
    code-fold: false
    toc: true
    toc-depth: 2
    reference-location: margin
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
    echo: true
    fig-width: 7.5
    fig-height: 4.5
    fig-align: center
---

```{r}
#| label: setup
#| results: hide
#| message: false
library(rFAMS)
library(ggplot2)  ## to make figures
library(tidyr)    ## for pivot_longer
library(dplyr)    ## for filter
library(metR)     ## geom_contour2
```

The objective of this article is to demonstrate how to use rFAMS to calculate yield based on multiple minimum length limit with multiple values of conditional fishing and conditional natural mortality.

# Build a life history parameter list
The first step for using any of the rFAMS simulation models is to create an object with life history parameters using the `makeLH()` function. THe `makeLH()` function creates a list of the life history parameters needed. Including, the initial number of recruits, maximum age of fish in your population, von Bertalanffy growth model parameters ($L_\infty$, $K$, and $t_0$), and parameters from the log10-transformed weight length model (alpha and beta). By default, the `makeLH()` returns a list. These values can be generated from functions in the [FSA package](https://fishr-core-team.github.io/FSA/index.html). The example below uses the following life history parameters:
- NO is the initial number of recruits, set to 100.
- tmax is the maximum age in the population in years, set to 15.
- Linf is the point estimate of asymptotic mean length from the von Bertalanffy growth model, set to 592mm.
- K (upper case) is the point estimate of the Brody growth coefficient from the von Bertalanffy growth model, set to 0.20.
- t0 is the point estimate of the x-intercept (i.e., theoretical age at a mean length of 0) from the von Bertalanffy growth model, set to -0.3.
- LWalpha is the point estimate of alpha from the length-weight regression on the log10 scale, set to -5.528.
- LWbeta is the point estimate of beta from the length-weight regression on the log10 scale, set to 3.273.

```{r}
#| label: Create life history object
# create life history parameter object
LH <- makeLH(N0=100,tmax=15,Linf=592,K=0.20,t0=-0.3,LWalpha=-5.528,LWbeta=3.273)
```

# Estimate yield for multiple minimum length limit and multiple conditional fishing and conditional natural mortality.

The function `yprBH_minLL_var()` function is used when yield is estimated with a single fixed minimum length limit. This function requires the minimum length `minLL`; specified range of conditional fishing mortality by setting the minimum `cfmin`, maximum `cfmax`, and increment `cfinc`; specified range of conditional natural mortality by setting the minimum `cmmin`, maximum `cmmax`, and increment `cminc`; set any length of interest to monitor `loi`; and the life history parameters `lhparams`.

rFAMS includes a function `est_natmort()` to estimate instantaneous natural mortality (M) and conditional natural mortality (cm) using parameters specified in the life history parameter object. See the [FSA package](https://fishr-core-team.github.io/FSA/index.html) for other methods of calculating M and cm. To generate the range and average M and cm using rFAMS:

```{r}
#| label: estimate natural mortality
est_natmort(LH, incl.avg = TRUE)
```

Once you have decided the range of cf and cm and decided what minimum length limit to consider, you can now proceed to estimating yield. The following example uses the life history object created above with a minimum length limit of 200mm, cf from 0.1 to 0.9 with increments of 0.1, cm from 0.1 to 0.9 with increments of 0.1, monitors lengths of 200, 250, 300, and 350mm.

```{r}
#| label: estimate yield
Res_1 <- yprBH_minLL_var(lengthmin=200,lengthmax=500,lengthinc=50, #creates minimum length limits
                         cfmin=0.1,cfmax=0.9,cfinc=0.1, #creates vector of cf values
                         cmmin=0.1,cmmax=0.9,cminc=0.1, #creates vector of cm values
                         loi=c(400, 450, 500, 550),     #sets lengths of interest to monitor
                         lhparms=LH)                    #Specifies life history parameters
```

The output object will be a data.frame with the following calculated values:

- yield is the estimated yield (in g).
- exploitation is the exploitation rate.
- Nharvest is the number of harvested fish.
- Ndie is the number of fish that die of natural deaths.
- Nt is the number of fish at time tr (time they become harvestable size).
- avgwt is the average weight of fish harvested.
- avglen is the average length of fish harvested.
- tr is the time for a fish to recruit to a minimum length limit (i.e., time to enter fishery).
- F is the instantaneous rate of fishing mortality.
- M is the instantaneous rate of natural mortality.
- Z is the instantaneous rate of total mortality.
- S is the (total) annual rate of survival.
- N at xxx mm is the number that reach the length of interest supplied. There will be one column for each length of interest.

View the first few rows of the output

```{r}
#| label: view results
head(Res_1)
```

# Plot results

We will now create a series of figures to aid in interpreting the output. First, a custom theme is created to use across all plots.

```{r}
#| label: create custom plot
# Custom theme for plots (to make look nice)
theme_FAMS <- function(...) {
 theme_bw() +
 theme(
   panel.grid.major=element_blank(),panel.grid.minor=element_blank(),
   axis.text=element_text(size=14,color="black"),
   axis.title=element_text(size=16,color="black"),
   axis.title.y=element_text(angle=90),
   axis.line=element_line(color="black"),
   panel.border=element_blank()
 )
}
```

## Yield as a function of exploitation

The first graph will be a yield curve that displays the relationship between yield as a function of exploitation for a specified conditional natural morality `cm`. The example below uses `cm` = 0.30, which is close to the mean estimated `cm` above of 0.27

```{r}
#| label: generate yield plot
# Round cm to account for "floating point arithmetic inaccuracies"
# will be fixed in next version of rFAMS so this isn't needed
Res_1$cm <- round(Res_1$cm,8)

# Extract results for cm=0.40 and minimum length limit=400
plot_dat <- Res_1 |> dplyr::filter(cm==0.30,minLL==400)

ggplot(data=plot_dat,mapping=aes(x=u,y=yield)) +
  geom_point() +
  geom_line() +
  labs(y="Total yield (g)",x="Exploitation (u)") +
  theme_FAMS()
```

This next graph plots yield curves as a function of exploitation with multiple minimum lengths for a specified conditional natural morality `cm`. The example below uses `cm` = 0.30, as above 

```{r}
#| label: generate yield plot multiple MLL
# Round cm to account for "floating point arithmetic inaccuracies"
# will be fixed in next version of rFAMS so this isn't needed
Res_1$cm <- round(Res_1$cm,8)

# Extract results for cm=0.40 and minimum length limit=400
plot_dat <- Res_1 |> dplyr::filter(cm==0.30)

ggplot(data=plot_dat,mapping=aes(y=yield,x=u,group=minLL,color=minLL)) +
  geom_line(linewidth=1) +
  scale_color_gradient2(high="black") +
  labs(y="Total yield (g)",x="Exploitation (u)",color="Min Length Limit") +
  theme_FAMS()

```

## Number of fish that reach a specified size as a function of exploitation

The next graph demonstrates how to explore the number of fish reaching a specified length. This figure creates a plot showing the number of fish reaching 450mm as a function of exploitation. Note this is based on a conditional natural mortality `cm` of 0.30 which was filtered out in the code block above.

```{r}
#| label: Plot of fish reaching 450mm
# Extract results for cm=0.40 and minimum length limit=400
plot_dat <- Res_1 |> dplyr::filter(cm==0.40,minLL==400)

ggplot(data=plot_dat,mapping=aes(x=u,y=`N at 450 mm`)) +
  geom_point() +
  geom_line() +
  labs(y="Number of fish at 300 mm",x="Exploitation (u)") +
  theme_FAMS()
```

The next example graph plots the number of fish reaching all the lengths of interests we monitored as a function of exploitation with conditional natural mortality `cm` of 0.30 which was filtered out above.

```{r}
#| label: Plot of fish reaching all sizes
# Select columns for plotting and convert to long
plot_data_long <- plot_dat %>%
 select(u,`N at 400 mm`, `N at 450 mm`, `N at 500 mm`, `N at 550 mm`) %>%
 pivot_longer(!u, names_to="loi",values_to="number")

# Generate plot
ggplot(data=plot_data_long,mapping=aes(x=u,y=number,group=loi,color=loi)) +
 geom_point() +
 scale_color_discrete(name="Yield",labels=c("N at 400 mm", "N at 450 mm", "N at 500 mm", "N at 550 mm"))+
 geom_line() +
 labs(y="Number of fish",x="Exploitation (u)") +
 theme_FAMS() +
 theme(legend.position = "top")+
 guides(color=guide_legend(title="Length of interest"))
```

## Isopleths of yield and number harvested

The final two graphs demonstrate how to construct isopleths using the `ggplot2` and `metR` packages. We will use the same data from above that is filtered for a conditional natural mortality `cm` of 0.30.


```{r}
#| label: Isoploths
# Extract results for cm=0.40 and minimum length limit=400
plot_dat <- Res_1 |> dplyr::filter(cm==0.30)

# Yield isopleths for varying minLL and exploitation with cm=0.40
# Using same data as previous example
ggplot(data=plot_dat,mapping=aes(x=u,y=minLL,z=yield)) +
  geom_contour2(aes(label = after_stat(level))) +
  xlab("Exploitation (u)") +
  ylab("Minimum length limit (mm)") +
  theme_FAMS()

# Same isopleth as previous but using number harvested instead of yield
ggplot(data=plot_dat,mapping=aes(x=u,y=minLL,z=Nharvest)) +
  geom_contour2(aes(label = after_stat(level))) +
  xlab("Exploitation (u)")+
  ylab("Minimum length limit (mm)")+
  theme_FAMS()
```
